Parsing Postfix log files

### 2009-02-16 doc/logparser-architecture.tex
Renamed labels in architecture and implementation chapters and fixed references. Improvements to Makefile, and added new helper targets. Some small improvements to architecture chapter.

@@

It should be clear from the earlier Postfix background (\sectionref{postfix
background}) that log files produced by Postfix vary widely from host to
host, depending on the set of restrictions chosen by the administrator.
With this in mind, one of the parser's design aims was to make adding new
rules as easy as possible, to enable administrators to properly parse their
own log files.  To enable this the architecture is divided into three
parts:

XXX UPDATE SECTION REFERENCES BELOW\@: THEY NEED TO REFER TO BOTH THE
ARCHITECTURE AND THE IMPLEMENTATION\@.

\begin{description}

    \item [framework] The framework is responsible for loading rules,
        managing data structures, finding the rule that matches each log
        line, invoking the correct action, etc\@.  The framework is
        described in detail in \sectionref{framework}.  XXX EXTEND
        FRAMEWORK SECTION IF NECESSARY\@.

    \item [rules] Rules match individual log lines and determine which
        actions will be executed.  Rules provide an easily extensible
        method of associating log lines with actions, and are described in
        detail in \sectionref{rules}.

    \item [actions] Actions are invoked to deal with a log line once it has
        been identified by the rules: actions modify data structures,
        handle complications, and cause data to be saved to the database.
        The actions do the work of reconstructing the journey a mail takes
        through Postfix.  Details of the actions available in the Postfix
        parser can be found in \sectionref{actions-in-detail}, and
        \sectionref{adding new actions} describes the process of adding new
        actions.

\end{description}

@@

Separating the rules from the actions and framework makes it possible to
parse new log lines without modifying the core parsing algorithm.  Adding a
new rule with the action to invoke and a regex to match the log lines is
trivial in comparison to understanding a program's entire parsing
algorithm, identifying the correct location to change and making the
appropriate changes.  Bear in mind that the changes must be made without
adversely affecting existing parsing, particularly as there may be edge
cases that are not immediately obvious ---
\sectionref{yet-more-aborted-delivery-attempts} describes a complication
that occurs only four times in \numberOFlogFILES{} log files tested.
Requiring changes to the parser's code also complicates upgrades, as the
changes must be preserved during the upgrade, and may clash with changes
made by the developer (\sectionref{adding new actions} describes safely
adding new actions).  \parsername{} allows the user to add new rules to the
database without changing the parsing algorithm, unless the new log lines
to be parsed require functionality not already provided by the algorithm.
If the new log lines do require new functionality, new actions can be added
to the parser without modifying existing actions or other parts of the
algorithm; only in the rare event that the new actions require support from
other sections of the code will more extensive changes be required.

@@

\label{rule in architecture}

@@

\label{rule characteristics}

@@

\label{overlapping rules}

@@

\label{framework}

>>>>

XXX WHEN FINISHED, CHECK ALL REFERENCES TO ENSURE THEY POINT AT THE CORRECT
CHAPTER\@.

@@

It should be clear from the earlier Postfix background (\sectionref{postfix
background}) that log files produced by Postfix vary widely from host to
host, depending on the set of restrictions chosen by the administrator.
With this in mind, one of the parser's design aims was to make adding new
rules as easy as possible, to enable administrators to properly parse their
own log files.  To enable this the architecture is divided into three
parts: framework, actions and rules.  Each will be discussed separately,
but first an overview:

XXX UPDATE SECTION REFERENCES BELOW\@: THEY SHOULD ONLY REFER TO THE
ARCHITECTURE, THE IMPLEMENTATION CAN REFER BACK TO HERE\@.

\begin{eqlist}

    \item [Framework]  The framework is the structure that actions and
        rules plug into.  It provides the parsing loop, shared data
        storage, loading and validation of rules, storage of results, and
        other support functions.  The framework is described in detail in
        \sectionref{framework in architecture}.

    \item [Actions] Each action performs the work required to deal with a
        single category of inputs, e.g.\ processing data from rejections.
        Actions are invoked to deal with a log line once it has been
        identified by the rules: actions modify data structures, handle
        complications, and cause data to be saved to the database.

    \item [Rules]  The rules are responsible for classifying inputs,
        specifying the action to invoke and the regex that matches the
        inputs and extracts data.  Rules provide an easily extensible
        method of associating log lines with actions, and are described in
        detail in \sectionref{rules in architecture}.

\end{eqlist}

@@

Separating the rules from the actions and framework makes it possible to
parse new log lines without modifying the core parsing algorithm.  Adding a
new rule with the action to invoke and a regex to match the log lines is
trivial in comparison to understanding a program's entire parsing
algorithm, identifying the correct location to change and making the
appropriate changes.  Bear in mind that the changes must be made without
adversely affecting existing parsing, particularly as there may be edge
cases that are not immediately obvious ---
\sectionref{yet-more-aborted-delivery-attempts} describes a complication
that occurs only four times in \numberOFlogFILES{} log files tested.
Requiring changes to the parser's code also complicates upgrades, as the
changes must be preserved during the upgrade, and may clash with changes
made by the developer (\sectionref{adding new actions in implementation}
describes safely adding new actions).  \parsername{} allows the user to add
new rules to the database without changing the parsing algorithm, unless
the new log lines to be parsed require functionality not already provided
by the algorithm.  If the new log lines do require new functionality, new
actions can be added to the parser without modifying existing actions or
other parts of the algorithm; only in the rare event that the new actions
require support from other sections of the code will more extensive changes
be required.

@@

\label{rules in architecture}

@@

\label{rule characteristics in architecture}

@@

\label{overlapping rules in architecture}

@@

XXX EXTEND FRAMEWORK SECTION IF NECESSARY\@.

\label{framework in architecture}

### 2009-02-16 doc/logparser-background.tex
Renamed labels in architecture and implementation chapters and fixed references. Improvements to Makefile, and added new helper targets. Some small improvements to architecture chapter.

Postfix's design follows the Unix philosophy of \textit{``Write programs
that do one thing and do it well''\/}~\cite{unix-philosophy}, and it is
separated into multiple programs that each perform one of the tasks
required of an \acronym{MTA}\@: receive mail, send mail, local delivery of
mail, etc.\ (full details can be found in~\cite{postfix-overview}).  Each
log line contains the name of the Postfix component that produced it, and
this information is used when determining which rules should be used to
parse each log line (see \sectionref{rule characteristics} for details).
Postfix's design is strongly influenced by security concerns: those
components that interact with other hosts are not
privileged,\footnote{Privilege means the power to perform actions that are
limited to the administrator, and not available to ordinary users.} so bugs
in those components will not give an attacker extra privileges; those
components that are privileged do not interact with other hosts, making it
much more difficult for an attacker to exploit any bugs that may exist in
those components.

>>>>

Postfix's design follows the Unix philosophy of \textit{``Write programs
that do one thing and do it well''\/}~\cite{unix-philosophy}, and it is
separated into multiple programs that each perform one of the tasks
required of an \acronym{MTA}\@: receive mail, send mail, local delivery of
mail, etc.\ (full details can be found in~\cite{postfix-overview}).  Each
log line contains the name of the Postfix component that produced it, and
this information is used when determining which rules should be used to
parse each log line (see \sectionref{rule characteristics in architecture}
for details).  Postfix's design is strongly influenced by security
concerns: those components that interact with other hosts are not
privileged,\footnote{Privilege means the power to perform actions that are
limited to the administrator, and not available to ordinary users.} so bugs
in those components will not give an attacker extra privileges; those
components that are privileged do not interact with other hosts, making it
much more difficult for an attacker to exploit any bugs that may exist in
those components.

### 2009-02-16 doc/logparser-conclusion.tex
Renamed labels in architecture and implementation chapters and fixed references. Improvements to Makefile, and added new helper targets. Some small improvements to architecture chapter.

Given the diverse nature of the log lines and the ease with which
administrators can cause new log lines to be logged (\sectionref{postfix
background}), enabling users to easily extend the parser to deal with new
log lines is a design imperative (\sectionref{parser design}).  Despite the
resulting initial increase in complexity the task is quite tractable,
though it does raise efficiency and optimisation questions (answered in
\sectionref{parser efficiency}).  Providing a tool to ease the generation
of regexes from unparsed log lines should greatly help users add rules
to parse previously unparsed log lines (\sectionref{creating new rules}).

The architecture not only substantially eases the parsing of new log
lines, it makes adding new actions (\sectionref{adding new actions}) a
relatively simple task.  The simplicity of adding a new action frees the
implementor from worrying about how their action might interfere with
existing parsing, so they can focus on correctly implementing their new
action.

@@

The real difficulties arise once the parser is successfully dealing with
90\% of the log lines, as the irregularities and complications explained in
\sectionref{complications} only become apparent once the vast majority of
log lines have been parsed successfully.  Adding new rules to deal with
numerous, infrequently occurring log lines is a simple, albeit tiresome,
task, (though alleviated by the tool described in \sectionref{adding new
actions}), whereas dealing with mails where information appears to be
missing is much more grueling.  Trawling through the log files, looking for
something out of the ordinary, possibly hundreds or even thousands of log
lines away from the last mention of the queueid in question, is extremely
time consuming and error prone.  Sometimes the task is not to spot the line
that is unusual, but to spot that a line that is normally present is
missing, i.e.\ to realise that one line amongst thousands is absent.  In
all cases the evidence must be used to construct a hypothesis to explain
the irregularities, and that hypothesis must then be tested in the parser;
if successful, the parser must be modified to deal with the irregularities,
without adversely affecting the existing functionality.  The complications
described in \sectionref{complications} were solved in the order they are
described in, and that order closely resembles the frequency with which
they occur; the most frequently occurring complications dominate the
warning messages produced, and so naturally they are the first
complications to be dealt with.

>>>>

Given the diverse nature of the log lines and the ease with which
administrators can cause new log lines to be logged (\sectionref{postfix
background}), enabling users to easily extend the parser to deal with new
log lines is a design imperative (\sectionref{parser design}).  Despite the
resulting initial increase in complexity the task is quite tractable,
though it does raise efficiency and optimisation questions (answered in
\sectionref{parser efficiency}).  Providing a tool to ease the generation
of regexes from unparsed log lines should greatly help users add rules
to parse previously unparsed log lines (\sectionref{creating new rules in
implementation}).

The architecture not only substantially eases the parsing of new log lines,
it makes adding new actions (\sectionref{adding new actions in
implementation}) a relatively simple task.  The simplicity of adding a new
action frees the implementor from worrying about how their action might
interfere with existing parsing, so they can focus on correctly
implementing their new action.

@@

The real difficulties arise once the parser is successfully dealing with
90\% of the log lines, as the irregularities and complications explained in
\sectionref{complications} only become apparent once the vast majority of
log lines have been parsed successfully.  Adding new rules to deal with
numerous, infrequently occurring log lines is a simple, albeit tiresome,
task, (though alleviated by the tool described in \sectionref{adding new
actions in implementation}), whereas dealing with mails where information
appears to be missing is much more grueling.  Trawling through the log
files, looking for something out of the ordinary, possibly hundreds or even
thousands of log lines away from the last mention of the queueid in
question, is extremely time consuming and error prone.  Sometimes the task
is not to spot the line that is unusual, but to spot that a line that is
normally present is missing, i.e.\ to realise that one line amongst
thousands is absent.  In all cases the evidence must be used to construct a
hypothesis to explain the irregularities, and that hypothesis must then be
tested in the parser; if successful, the parser must be modified to deal
with the irregularities, without adversely affecting the existing
functionality.  The complications described in \sectionref{complications}
were solved in the order they are described in, and that order closely
resembles the frequency with which they occur; the most frequently
occurring complications dominate the warning messages produced, and so
naturally they are the first complications to be dealt with.

### 2009-02-16 doc/logparser-implementation.tex
Renamed labels in architecture and implementation chapters and fixed references. Improvements to Makefile, and added new helper targets. Some small improvements to architecture chapter.

The database schema can be conceptually divided in two: the rules used to
parse log files, and the data saved from the parsing of log files.  Rules
have the fields required to parse the log lines, extract data to be saved,
and specify the action to be executed; they also have several fields that
aid the user in understanding the meaning of the log lines parsed by each
rule.  The rules are described in detail in \sectionref{rules} but the
fields are covered in \sectionref{rule attributes}.

@@

Rules are discussed in detail in \sectionref{rules}, but the structure of
the rules table is covered here.  Rules are created by the user, not the
parser, and will not be modified by the parser (except for the hits and
hits\_total fields).  Rules parse the individual log lines, extracting data
to be saved in the connections and results tables, and specifying the
action to take for that log line.

@@

    \item [postfix\_action] The action Postfix must have taken to generate
        this log line; the possible actions are described later in
        \sectionref{list of postfix actions}.

@@

    \item [action] The action that will be invoked when this rule matches a
        log line; a full list of actions and the parameters they are
        invoked with can be found in \sectionref{actions-in-detail}.

@@

    \item [priority] This is the user-configurable companion to hits: when
        the list of rules is sorted, priority overrides hits.  This allows
        more specific rules to take precedence over more general rules
        (described in \sectionref{overlapping rules}).

@@

\label{rules}

@@

\label{example rule}

@@

    \item [action:] The action to be executed if the regex matches
        successfully.  See \sectionref{actions-in-detail} for details of
        the actions implemented by the parser, \sectionref{actions in
        architecture} for the role of actions in the parser architecture.

@@

\label{creating new rules}

@@

@@

\label{actions-in-detail}

@@

\label{list of postfix actions}

@@

\label{adding new actions}

@@

\label{limitations-improvements}

>>>>

The database schema can be conceptually divided in two: the rules used to
parse log files, and the data saved from the parsing of log files.  Rules
have the fields required to parse the log lines, extract data to be saved,
and specify the action to be executed; they also have several fields that
aid the user in understanding the meaning of the log lines parsed by each
rule.  The rules are described in detail in \sectionref{rules in
implementation} but the fields are covered in \sectionref{rule attributes}.

@@

Rules are discussed in detail in \sectionref{rules in implementation}, but
the structure of the rules table is covered here.  Rules are created by the
user, not the parser, and will not be modified by the parser (except for
the hits and hits\_total fields).  Rules parse the individual log lines,
extracting data to be saved in the connections and results tables, and
specifying the action to take for that log line.

@@

    \item [postfix\_action] The action Postfix must have taken to generate
        this log line; the possible actions are described later in
        \sectionref{list of postfix actions in implementation}.

@@

    \item [action] The action that will be invoked when this rule matches a
        log line; a full list of actions and the parameters they are
        invoked with can be found in \sectionref{actions in detail in
        implementation}.

@@

    \item [priority] This is the user-configurable companion to hits: when
        the list of rules is sorted, priority overrides hits.  This allows
        more specific rules to take precedence over more general rules
        (described in \sectionref{overlapping rules in architecture}).

@@

\label{rules in implementation}

@@

\label{example rule in implementation}

@@

    \item [action:] The action to be executed if the regex matches
        successfully.  See \sectionref{actions in detail in implementation}
        for details of the actions implemented by the parser,
        \sectionref{actions in architecture} for the role of actions in the
        parser architecture.

@@

\label{creating new rules in implementation}

@@

The actions do the work of reconstructing the journey a mail takes through
Postfix.  Details of the actions available in the Postfix parser can be
found in \sectionref{actions in detail in implementation}, and
\sectionref{adding new actions in implementation} describes the process of
adding new actions.

@@

\label{actions in detail in implementation}

@@

\label{list of postfix actions in implementation}

@@

\label{adding new actions in implementation}

@@

\label{limitations and improvements in implementation}

### 2009-02-16 doc/logparser-introduction.tex
Fixed some 'overfull hbox' problems.
Renamed labels in architecture and implementation chapters and fixed references. Improvements to Makefile, and added new helper targets. Some small improvements to architecture chapter.

Section~\ref{background} provides background information useful in
understanding the thesis, parser, and architecture.  It introduces the idea
of using a database schema as an \acronym{API}, to provide a
language-neutral interface to the data gathered.  The novel separation of
rules, actions, and framework is discussed, giving the reasons that
approach was taken when designing the parser.

@@

Section~\ref{rules} discusses the parsing rules in detail, explaining the
purpose and usage of each field in a rule, referring to an example rule and
sample data it matches successfully against.  The pros and cons of
overlapping rules are considered, including techniques for detecting
unintentional overlaps.  Rule efficiency concerns are discussed, in
particular the optimisations used by the algorithm.  The section concludes
with a description of using the tools provided with the parser to generate
new rules (specifically the regex in each rule) from unparsed log lines.

@@

Section~\ref{limitations-improvements} lists the limitations of the
algorithm, then suggests some ways of dealing with them, with the goal of
improving parsing and reproduction of the journey a mail takes through
Postfix.

>>>>

Section~\ref{background} provides background information useful in
understanding the thesis, parser, and architecture.

@@

Section~\ref{rules in architecture} discusses the parsing rules in detail,
explaining the purpose and usage of each field in a rule, referring to an
example rule and sample data it matches successfully against.  The pros and
cons of overlapping rules are considered, including techniques for
detecting unintentional overlaps.  Rule efficiency concerns are discussed,
in particular the optimisations used by the algorithm.  The section
concludes with a description of using the tools provided with the parser to
generate new rules (specifically the regex in each rule) from unparsed log
lines.

@@

Section~\ref{limitations and improvements in implementation} lists the
limitations of the algorithm, then suggests some ways of dealing with them,
with the goal of improving parsing and reproduction of the journey a mail
takes through Postfix.
